Information Distribution Depends on Language-Specific Features

### 2019-02-01 cogsci/position-info.Rmd
through E3

abstract: >
    We argue for a reformulation of the Uniform Information Density hypothesis @levy2007 in which the grammatical, phonological and morphological features of a language determine how information is distributed throughout utterances. We expand on a method from @yu2016 to compute entropy based on word position within utterances as a derivation of information distribution. We analyze adult and child spoken corpora as well as text corpora drawn from a variety of languages on Wikipedia. We find that each language has a robust, characteristic entropy curve across word position, and present evidence that these entropy curves are in part determined by typological features. We conclude that speakers may tend towards communicating information uniformly, but the distribution of information in their utterances is conditioned on the language they communicate in.

keywords: >
    Entropy; information; information theory; communication

@@

# Introduction

@@

<!-- **Evidence for UID here. Maybe see if you can present this work at a slightly higher level** -->
The original evidence in @levy2007 finds that the insertion of complementizers in relative clauses in English corresponds to where neighboring words have high information content. Similarly, @frank2008 argues that contradictions in English such as "you're" do not occur when neighboring words are highly informative. The evidence in favor of UID largely been situation-specific and English-language driven, while the hypothesis itself has been applied broadly over the past decade. Applications include determining whether linguistic alignment takes place [@jaeger2013], Zipfian word length distributions [@piantadosi2011], communication efficiency [@mahowald2013], dialogue and turn-taking [@xu2018] and the significance of ambiguity in language [@piantadosi2012], among other research.

<!-- **Evidence against UID here. Again, can you present this at a higher level?** -->
However, other recent work has contradicted the UID hypothesis. Similar to the original @levy2007, @zhan2018 focuses on information distribution at particular points in sentences. @zhan2018 finds that more information-rich classifiers in Mandarin Chinese are produced when production of the neighboring noun is difficult, not when the information content is high. @jain2018 examine word order across spoken sentences in Hindi, a freer word order language than English, and find that information density has no significant effect on word order.

Recently, @yu2016 developed a more direct test of the Uniform Information Density hypothesis, applying the logic used by @genzel2002 to look at the distribution of information *within* individual sentences. <!-- **SOME HIGHER LEVEL DESCRIPTION HERE, WE'll GET INTO THE DETAILS BELOW** -->Because people process language incrementally--using the previous words in a sentence to predict the words that will come next--the amount of information that a word contains when seen in isolation should increase over the course of a sentence. Analyzing a large corpus of written English, they find a different pattern: Entropy increases over the first few words of an utterance and then remains constant until the final word where it again jumps up (see Figure \ref{fig:our_replication}). @yu2016 conclude that the Uniform Information Density hypothesis must not hold for medial words in a sentence.

We extend and generalize @yu2016 in three ways: we confirm that this same pattern is found in spoken English--in both adult and child speakers. We then examine entropy curves cross-linguistically, finding a diversity of shapes across the world's languages. Finally, we show that entropy curves are predictable from the structure of individual languages, namely word order. Taken together, our results suggest a refinement of the Uniform Information Density hypothesis: speakers may structure their utterances to optimize information density, but they must do so under the constraints of the language they use to communicate.

The UID hypothesis predicts that information transmission rate will tend towards uniformity in both speech and text. We begin by examining speech to determine if we obtain the entropy curve shape we expect for spoken communication. We begin with the same source as @levy2007 and @jaeger2010, the Switchboard corpus of adult telephone conversations. We also examine child-adult conversations in the CHILDES TalkBank [@demuth2006; @macwhinney2014] corpora database of spoken adult-child conversations in multiple languages. Children are not fully developed speakers, so we want to compare the entropy curve we obtain by computing over the utterances in the CHILDES corpora to the utterances in the adult-adult Switchboard corpus. Finally, we extend our cross-linguistic findings on a small number of languages from the CHILDES experiment to a direct comparison of several dozen languages in Wikipedia.

## General methods

Here we will describe our adaptation of the by-word entropy method in @yu2016, which uses the text portion of the British National Corpus [@clear1993]. We replicate their results at the end of this subsection. Given a text or speech corpus divided into individual utterances, we partition the corpus by utterance length in number of words. For each word position $X$ of utterances of length $k$, we define $w$ as a unique word occurring in position $X$. We further define $p(w)$ as the number of times word $w$ occurs in position $X$, divided by the number of total words that occur in position $X$ i.e. the number of sentences of length $k$. This creates a probability distribution over the words occurring in position $X$, and computing the Shannon entropy $H(X)$ @shannon1948 of this probability distribution gives the positional entropy of position $X$ in utterances of length $k$.

@@

With this measure, we compute the unigram entropy at each position of sentences of each length within the corpus. The result of this method can be plotted for each utterance length as an *entropy curve*, which can be visually compared across utterance length to observe the how the unigram entropy changes across absolute positions in each of the utterances. @genzel2002 similarly examine a unigram entropy measure on sentences, and found that entropy at the sentence level increases linearly with sentence index within a corpus. UID applies this uniformity of entropy rate in sentences to all levels of speech, and so our method obtained from @yu2016, which examines text at the word level, should find an affine function at the word level.

The entropy curves capture individual variation across positions in utterances of the same length. This allows us to directly observe and judge the amount of variation in words that appear in an individual position of a sentence. We can directly compare any two positions within utterances to determine the amount of uncertainty, and therefore information, on average contained by words within that position of utterances. We are applying the same approach as in @genzel2002, but within sentences instead of across sentences.

The positional entropy at the beginning of sentences is low, then rises and plateaus for sentence-medial word positions before dropping slightly in the second-to-last position and rising again. However, this curve could simply be a result of the written words in the BNC. We will expand this analysis and replicate the result with a spoken American English corpus.

# Experiment 1

@@

We replicated the @yu2016 results and found the same characteristic three-step distribution as they did for the BNC entropy curves.

We first used the same source for data as @levy2007, the Switchboard corpus of American telephone conversations @godfrey1992. The Switchboard corpus contains telephone conversations of American English between strangers, with individual utterances ranging from one word to several dozen words. We divided the Switchboard corpus by utterance length and computed the positional entropy for each position within each sentence length, then plotted the result as an entropy curve.

Our plots for three representative utterance lengths are below. We found the same characteristic shape for the American English Switchboard corpus as in our replication of the @yu2016 results for the BNC.

```{r read_and_plot_switchboard, fig.env = "figure*", fig.width=7, fig.height=4, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom). Points average entropies, error bars show 95\\% confidence intervals computed by non-parametric bootstrap."}

@@

The shape we find in these two English data is notable for a number of reasons, but what we find most interesting is one, the shape of the distribution does not follow our predictions from UID and two, the distribution is robust across an unrelated pair of spoken and written English corpora. This suggests that the entropy curve is characteristic of the English language at least, and is not a specific effect of computing positional entropies in the written British English texts in the BNC. For our next step, we want to expand the positional entropy analysis to child-adult conversational corpora and languages beyond English to see if the same three-step distribution appears.

# Experiment 2

After replicating the @yu2016 results using the spoken conversation corpus from @levy2007, we want to determine whether the entropy curves will remain robust over child-adult conversational corpora. We also wanted to see whether the curves would show the same characteristic three-step shape across languages. To accomplish both of these goals, we used data from the CHILDES Talkbank system @macwhinney2014.

The corpora in the CHILDES database consists of short and often disconnected utterances across hours of recordings, the unigram entropy measure is unaffected by context or lack thereof in utterances by the adults and children in the corpora. For our first corpus, we used the Providence English corpus @demuth2006. The Providence corpus recorded interactions between children between 1 and 3 years old and their parents in the home. We divided the corpus by speaker into child and non-child categories. We further divided the corpus by utterance length, so that all sentences of length $k$ (e.g. $6$) were grouped together. Finally, within each utterance length, we computed the unigram entropy measure for each position. Computation of positional entropy was identical to our general method after dividing for utterance length.

We also ran this analysis on Spanish and Mandarin corpora from CHILDES. Spanish is an Indo-European language like English, possessing similar grammar, word order and numerous cognate word. Mandarin Chinese is typologically completed unrelated to English. We used the Shiro corpus for Spanish @shiro2000, which contains prompted narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. We used the Zhou dinner corpus for Mandarin Chinese @li2015, which contains dinner conversations between 5 to 6-year-old children and their parents collected in Shanghai.

For each corpus, we accessed transcripts of the corpus provided through the TalkBank system and computed over the Latin alphabet transcriptions or transliterations of the original transcriptions. For Mandarin, we used pinyin transliterations of the utterances in the corpus with demarcated word boundaries. The Chinese characters used for writing Mandarin do not normally demarcate word boundaries by spacing words apart, and for normal Chinese writing including spaces between word boundaries can have a negative effect on reading times @bai2008.

We expect that our entropy analysis on the Providence corpus will produce an identical entropy curve shape to the BNC and Switchboard curves. We expect similar results to English in the Spanish Shiro entropy curve, while we expect that our Mandarin corpus will produce a different entropy curve shape.

@@

```{r plot_childes, fig.env = "figure*", fig.width=7, fig.height=4, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Representative Entropy Curves for Three Childes Corpora in English, Mandarin, and Spanish. Points average entropies, error bars show 95\\% confidence intervals computed by non-parametric bootstrap."}
childes_entropies <- read_csv(here("data/childes_entropies.csv"))

@@

The adult and child entropy curves track one another almost identically. This is surprising because UID would predict that the young children in the corpora who are not fully developed speakers would have more noise in their distribution of information rates. This not only indicates a robustness in the unigram entropy curve across speakers, but also across ages and addressees. We also observed a robustness across corpora for the same languages, and a robustness across utterance lengths within the same corpus's entropy curve. This shows that the concept of an "entropy curve" for a specific language is well-founded when considering speech data.

We found a distinct three-step distribution for English and Spanish CHILDES corpora, with a slight dip in the penultimate position of each sentence. The Mandarin corpus entropy curve, by comparison, has a noticeably lower positional entropy values in utterance-final positions than in utterance-penultimate positions. We attribute the penultimate dip in the English and Spanish entropy curves to the fact that most of the utterances in the CHILDES English and Spanish corpora we examined had a determiner such as "the" or "a" in the second-to-last position of utterances. The beginnings of utterances in the English and Spanish CHILDES data were usually pronouns or grammatical subjects, while the final words were grammatical objects and had a great deal of variation in the exact word that appeared in the utterance-final position.

These entropy curves are not what would be expected from UID. The robustness of the three-step distribution for English and its replication in Spanish do not resemble the affine function we would expect from UID. The Mandarin entropy curve, which does not at all resemble either the English/Spanish distribution or the predictions of UID, suggests that the entropy curve can vary from language to language. UID predicts that each language should have a similar distribution, which should approximate an affine function.

We found that all of the English corpora followed the same three-step entropy curve shape, regardless of whether the corpus contained spoken or written data. The Spanish corpus we used also followed this same shape, which we expected. However, Mandarin did not have the same shape, indicating that the shape of the entropy curves may arise from cross-linguistic variation. Similar grammatical and other features between two languages, such as Spanish and English, may produce a similar entropy curve.

>>>>

abstract: >
    Although languages vary widely in their structure, all are vehicles for the transmission of information. Consequently, aspects of speakers' word choice can be understood through the lens models of optimal communication. One prediction of these models is that speakers should keep the density of information constant over the duration of an utterance [@levy2007]. However, different languages have different coding models that constraint the space of speaker's choices (e.g. canonical word order). We  build on a mdethod developed by @yu2016 to analye the entropy curves of natural language productions across a diverse set of languages and in diverse written and spoken contexts. We show languages have characteristic constraints that they impose on speaker's choices that predict deviations from Uniform Information Density, and that cross-linguistic variability in these deviations is predictable in part from syntactic properties of those languages.

keywords: >
    Uniform information density; language structure; corpus analysis

@@

@@

<!-- **Evidence for UID here. Maybe see if you can present this work at a slightly higher level** -->
The original evidence in @levy2007 finds that the insertion of complementizers (e.g. "that") in relative clauses in English corresponds to where neighboring words have high information content. Similarly, @frank2008 argues that contradictions in English such as "you're" do not occur when neighboring words are highly informative. The evidence in favor of Uniform Information Density largely been situation-specific and English-language driven, while the hypothesis itself has been applied broadly over the past decade. Applications include determining whether linguistic alignment takes place [@jaeger2013], Zipfian word length distributions [@piantadosi2011], communication efficiency [@mahowald2013], dialogue and turn-taking [@xu2018] and the significance of ambiguity in language [@piantadosi2012], among other research.

<!-- **Evidence against UID here. Again, can you present this at a higher level?** -->
However, other recent work has contradicted the Uniform Information Density hypothesis. Similar to the original @levy2007, @zhan2018 focuses on information distribution at particular points in sentences. @zhan2018 finds that more information-rich classifiers in Mandarin Chinese are produced when production of the neighboring noun is difficult, not when the information content is high. @jain2018 examine word order across spoken sentences in Hindi, a freer word order language than English, and find that information density has no significant effect on word order.

Recently, @yu2016 developed a more direct test of the Uniform Information Density hypothesis, applying the logic used by @genzel2002 to look at the distribution of information *within* individual sentences. Because people process language incrementally--using the previous words in a sentence to predict the words that will come next--the amount of information that a word contains when seen in isolation should increase over the course of a sentence [@ferrer-i-cancho2013]. Analyzing a large corpus of written English, they find a different pattern: Entropy increases over the first few words of an utterance and then remains constant until the final word where it again jumps up (see top of Figure \ref{fig:read_and_plot_exp1}). @yu2016 conclude that the Uniform Information Density hypothesis must not hold for medial words in a sentence.

We extend and generalize @yu2016 in three ways: We confirm that this same pattern is found in spoken English--both in written language, and in conversational speech between adults and between parents and their children. Thus, this entropy curve is a robust feature of English productions. We then examine entropy curves cross-linguistically, and show that characteristic curves vary across langauge the world's languages. Finally, we show that this variation is predictable in part from the structure of individual languages (i.e. word order). Taken together, our results suggest a refinement of the Uniform Information Density hypothesis: speakers may structure their utterances to optimize information density, but they must do so under the predictable constraints of their language.

## Calculating entropy curves

In all oll of our studies, we used an adaptation of the by-word entropy method developed by  @yu2016.
Given a text or speech corpus divided into individual utterances, we partition the corpus by utterance length in number of words. For each word position $X$ of utterances of length $k$, we define $w$ as a unique word occurring in position $X$. We further define $p(w)$ as the number of times word $w$ occurs in position $X$, divided by the number of total words that occur in position $X$ i.e. the number of sentences of length $k$. This creates a probability distribution over the words occurring in position $X$, and computing the @shannon1948 entropy $H(X)$ of this probability distribution gives the positional entropy of position $X$ in utterances of length $k$.

@@

With this measure, we compute the unigram entropy at each position of sentences of each length within the corpus. The result of this method can be plotted for each utterance length as an *entropy curve*, which can be visually compared across utterance length to observe the how the unigram entropy changes across absolute positions in each of the utterances. @genzel2002 similarly examine a unigram entropy measure on sentences, and found that entropy at the sentence level increases linearly with sentence index within a corpus. Uniform Information Density applies this uniformity of entropy rate in sentences to all levels of speech, and so our method obtained from @yu2016, which examines text at the word level, should find a monotonically increasing  at the word level.

The entropy curves capture individual variation across positions in utterances of the same length. This allows us to directly observe and judge the amount of variation in words that appear in an individual position of a sentence. We can directly compare any two positions within utterances to determine the amount of uncertainty, and therefore information, on average contained by words within that position of utterances. This method is thus identical in logic to @genzel2002, but within sentences instead of across sentences.

# Study 1

We began by replicating Yu et al.'s [-@yu2016] analysis, computing entropy curves on the British National Corpus--a collection of predominantly written English [@clear1993]. We then applied the same method to the Switchboard corpus--a collection of spoken language [@godfrey1992]. This allows us to ask whether the characteristic function identified by @yu2016 is a general feature of English, or instead a function of written language.

# Data and Analysis

The British National Corpus consists of predominantly (90%) written language documents collected in the early 1980s and 1990s across a variety of genres (scientific articles, newspapers, fiction, etc). It also contains a small collection of spoken language. All together, it contains $\sim$ 100 million tokens. The Switchboard corpus is a collection of $\sim$ 2,400 telephone conversations between unacquainted adults prompted with subjects of conversation. Switchboard is the corpus used in Levy and Jaeger's [-@levy2007] original demonstration of the Uniform Information Density Hypothesis.

For each corpus, we computed entropy curves using the method described above for all sentences from length 4 to 30.

@@

# Results

The entropy curves computed in both the British National Corpus and Switchboard were remarkably consistent both across corpora and across the range of sentence lenghts we analyzed (Figure \ref{fig:read_and_plot_exp1}). Confirming Yu et al.'s [-@yu2016] findings, we find that
The positional entropy at the beginning of sentences is low, then rises and plateaus for sentence-medial word positions before dropping slightly in the second-to-last position and rising again.

```{r read_and_plot_exp1, fig.env = "figure*", fig.width=7, fig.height=4, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom). Points average entropies, error bars show 95\\% confidence intervals computed by non-parametric bootstrap."}

@@

The shape of positional entropy curves we find in these two corpora is notable for two reasons: (1) it does not follow our predictions from Uniform Information Density, and (2) the distribution is robust across written and spoken Engilsh. This suggests that the entropy curve is characteristic of the English language. We next asked whether this shape is a feature even of children's speech, and also whether it varies cross-linguistically.

# Study 2

To understand how robust this entropy curve is, we turned to conversational speech from parent-child interactions. If we find the same shape even in children's productions, we have even stronger evidence that the 3-step curve is a charcateristic feature of English. We thus turned to the Child Language Data Exchange System (CHILDES), a collection of transcripts of parent-child interactions [@macwhinney2014].

## Data and Analysis

We analyzed three seperate corpora in CHILDES: The Providence Corpus [@demuth2006], The Shiro corpus [@shiro2000], and the Zhou Dinner Corpus [@li2015]. The Providence corpus consists of conversations between six 1--3-year-old American English speaking children and their parents recorded in their homes. Providence corpus recorded interactions between children between 1 and 3 years old and their parents in the home. The Shiro Corpus consists of prompted Spanish-language narratives individually collected from over a hundred Venezualan schoolchildren, half from high SES backgrounds and half from low SES backgrounds. The Zhou Dinner Corpus contains dinner conversations between 5 to 6-year-old Mandarin speaking children and their parents collected in Shanghai. Spanish is an Indo-European language like English, possessing similar grammar, word order and numerous cognate word. In contrast, Mandarin Chinese is typologically completed unrelated to English.

We accessed the transcripts from each corpus using \texttt{childesr}, an R-interface to a database formatted version of CHILDES [@sanchez2019]. We divided all utterances from each transcript into those produced by target child, and those produced by all other speakers. We then applied the same entropy curve method described above. For Mandarin, we used pinyin transliterations of the utterances in the corpus with demarcated word boundaries. The Chinese characters used for writing Mandarin do not normally demarcate word boundaries by spacing words apart, and for normal Chinese writing including spaces between word boundaries can have a negative effect on reading times [@bai2008].

@@

```{r plot_childes, fig.env = "figure*", fig.width=6, fig.height=3, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Representative Entropy Curves for Three Childes Corpora in English, Mandarin, and Spanish. Points average entropies, error bars show 95\\% confidence intervals computed by non-parametric bootstrap."}
childes_entropies <- read_csv(here("data/childes_entropies.csv"))

@@

Across corpora, we found similar entropy curve shapes for adults and children, but distinct shapes for each language. Figure \ref{fig:plot_childes} shows representative curves across corpora, but shapes were robust across the full range utterances we analyzed. We found a distinct three-step distribution for English and Spanish CHILDES corpora, with a slight dip in the penultimate position of each sentence. The Mandarin corpus entropy curve, by comparison, has a noticeably lower positional entropy values in utterance-final positions than in utterance-penultimate positions.

These results present two important pieces of information. First, our analysis of the Providence corpus broadly replicates the shape we found in both the British National Corpus and in Switchboard. Thus, the entropy curve of English appears not just in adult-adult conversation, but even in speech produced by parents to their children, and speech produced by very young children to their parents. This suggests that it is a highly robust feature of English langauge, as it structures the productions of even pre-schooled aged children.

Second, the Entropy curves of English, Spanish, and Mandarin were not identical. None of these shapes resembled the monotonically increasing function predicted naively from Uniform Information Density, but also Mandarin was quite different from both English and Spanish. This suggests that the entropy curve can vary from language to language. One possibility is that this variation arises from typological features of languages, such as characteristic word order. We noticed, for instance, a relatively high density of determiners in the penultimate position of English and Spanish utterances, which could account for the penultimate dip in those languages. In our final analysis, we explored this possibility directly, analyzing a large set of Wikipedia corpora from diverse languages, and asking whether variability in their entropy curves was related to variability in their syntactic structure.
